Expanding tidy data principles to facilitate missing data exploration, visualization and assessment of imputations

### 2020-01-06 paper/tidy-missing-data-paper.Rmd
minor text changes

A data structure for missing data provides key benefits over on-the-fly creating a logical matrix where TRUE means missing, and FALSE means present. Firstly, it can be confusing to users whether TRUE or FALSE means missing or present; `NA` and `!NA` are clearer. Secondly, special missing values cannot be added fluently during analysis with a logical matrix. Finally, the logical matrix cannot capture which values are imputed, if imputation has already taken place. Imputing values on _nabular_ data automatically tracks these value changes.

Using additional columns to represent missingness information follows recently published best practices for data organization, described in @Ellis2017 and @Broman2017: (1) Keep one thing in a cell and (2) Describe additional features of variables in a second column. Here they suggest to indicate censored data with an extra variable called "VariableNameCensored", which would be TRUE if censored, otherwise FALSE. This information is now represented in the shadow columns as additional integers, along with the other missing information.

@@

The number of times certain variables go missing together can be visualized using an "upset plot" [@Conway2017], a visualization technique for showing the size and features of sets in data that is similar to a venn diagram, but scales better with more variables (Figure \@ref(fig:airquality-upset)).

@@

Figure \@ref(fig:airquality-upset) shows that only Ozone and Solar.R have missing values. The dots in the bottom right show the combinations of missingness, with the bottom left showing the missingness in each variable, and the top showing the missingness for the combinations. The bottom left shows that Ozone has the most missing values, and the top shows 2 cases where both Solar.R and Ozone have missing values together.

@@

To visualize the amount of missings in each variable, the data is transformed into _nabular_ form, then values are imputed values below the range of data using `impute_below` (by default imputing 10% below the range). Visualizing this as a histogram shows missing values on its left. Ozone is shown in a different color by referring to the variable `Ozone` as `Ozone_NA` in a fill aesthetic in ggplot, shown in Figure \@ref(fig:impute-shift-histogram).

>>>>

The _nabular_ data structure for missing data provides key benefits over on-the-fly creating a logical matrix where TRUE means missing, and FALSE means present. Firstly, it can be confusing to users whether TRUE or FALSE means missing or present; `NA` and `!NA` are clearer. Secondly, special missing values cannot be added fluently during analysis with a logical matrix. Finally, the logical matrix cannot capture which values are imputed, if imputation has already taken place. Imputing values on _nabular_ data automatically tracks these value changes.

Using additional columns to represent missingness information follows recently published best practices for data organization, described in @Ellis2017 and @Broman2017: (1) Keep one thing in a cell and (2) Describe additional features of variables in a second column. Here they suggest to indicate censored data with an extra variable called "VariableNameCensored", which would be TRUE if censored, otherwise FALSE. This information can now be represented in the shadow columns as special missings.

@@

The number of times certain variables go missing together can be visualized using an "upset plot" [@Conway2017], a visualization technique for showing the size and features of sets in data that is similar to a venn diagram, but scales better with more variables (Figure \@ref(fig:airquality-upset)). Figure \@ref(fig:airquality-upset) shows that only Ozone and Solar.R have missing values. The dots in the bottom right show the combinations of missingness, with the bottom left showing the missingness in each variable, and the top showing the missingness for the combinations. The bottom left shows that Ozone has the most missing values, and the top shows 2 cases where both Solar.R and Ozone have missing values together.

@@

@@

To visualize the amount of missings in each variable, the data is transformed into _nabular_ form, then values are imputed values below the range of data using `impute_below` (by default imputing 10% below the range). Visualizing this as a histogram shows the number of missing values on its left. Ozone is shown in a different color by referring to the variable `Ozone` as `Ozone_NA` in a fill aesthetic in ggplot, shown in Figure \@ref(fig:impute-shift-histogram).
